### 2016-08-04 latex/input/conclusion.tex
finished related work section of paper and added diagrams; began working on results section

We have demonstrated an effective and efficient computational pipeline for classifying patient phenotype based on metagenomic data. We have demonstrated that even relatively simple de novo assembly and clustering methods, when used within this pipeline, lead to significantly better performance results than the standard classifier used in the original Metagenome-Wide Association Study. Finally, we have shown the effectiveness of Multiple Instance Learning methods within metagenomics and phenotype prediction, particularly Distance-based Bag of Words methods. Future work could revolve around improving individual parts of the pipeline, such as better assembly and clustering methods, application of different multiple instance learning methods (other than D-BoW), and further attempts to generate more specific instance level labels and validate those labels against clinically verified labels.

>>>>

We have demonstrated an effective and efficient computational pipeline for classifying patient phenotype based on metagenomic data. We have demonstrated that even relatively simple de novo assembly and clustering methods, when used within this pipeline, lead to significantly better performance results than the standard classifier used in the original Metagenome-Wide Association Study. We have shown how to infer the most important OTUs in the disease pathology by using the SVM decision boundary and discussed the clinical importance of this ability. More generally, we have shown the effectiveness of Multiple Instance Learning methods within metagenomics and phenotype prediction, particularly Distance-based Bag of Words methods. Future work could revolve around improving individual parts of the pipeline, such as better assembly and clustering methods, application of different multiple instance learning methods (other than D-BoW), and further attempts to generate more specific instance level information and validate that information against clinical understanding of the diseases.

### 2016-08-04 latex/input/experiments.tex
finished related work section of paper and added diagrams; began working on results section

\subsection{Data, Software, and Hardware Used}

Our main dataset comes from a GWAS of Type 2 Diabetes (T2D) in Chinese adults, published by Qin et al \cite{qin041012}. The data is publicly available on the National Center for Biotechnology Information (NCBI) \cite{ncbi} website using the accession number SRP011011. The full dataset contains over 200 patients, but for our study we used a subset of 60 patients due to lengthy download and format conversion times. Of these 60 patients, 30 had T2D and 30 were controls. For the purpose of testing our cluster sampling methods, we also used a smaller dataset obtained from Dr. Patrick Gillevet of GMU, which is not publicly available. It contains 904 patients, 239 of whom have encephalopathy and 665 of whom are controls, but is much smaller because it uses 16S rRNA instead of whole genome data. The encephalopathy dataset contains about 1.3 million reads, while the T2D dataset contains about 1.3 billion reads.

As previously mentioned, we use the publicly-available UCLUST \cite{Edgar10} software for clustering. For classification, we use Support Vector Machines (SVMs) \cite{vap95}, one of the most powerful and versatile binary classifiers used 
in myriad applications. 
%
For a  binary classification problem, the SVM framework 
maximizes the separation between the two classes. In addition to the ubiquity and effectiveness of SVMs, they are well established in both supervised and multiple instance learning, making it easy to compare the two approaches. For our standard supervised SVM, we use svm-light \cite{joachims08}, a popular and publicly available implementation. For multiple instance learning, we use the python implementation by G. Doran and S. Ray ("MISVM") \cite{doran122013} for its ease of use and public availability.

\subsection{Efficient Unsupervised Clustering using UCLUST}
\label{uclustmethods}

From early experimental results, UCLUST can still be computationally infeasible for very large datasets such as the one we will be using, despite its generally excellent speed. For our T2D dataset, UCLUST was only able to cluster 2.8\% of the reads within 12 hours, at which point ARGO killed the batch job. UCLUST slows down as more clusters are formed, as each new read needs to be compared to more existing clusters. Thus, as the number of clusters grows, the execution time of UCLUST increases rapidly. The only way to feasibly complete the clustering phase in a reasonable amount of time was to cluster the patients individually in parallel, instead of clustering the reads of all patients together. However, this leads to the problem of having mismatched clusterings between patients, that is, cluster X from patient A may represent the same OTU as cluster Y from patient B. The solution is to compare the clusters from different patients and map them to each other, and then to combine clusters from different patients that represent the same OTU and output a single clustering output file for all patients. However, comparing every read from each cluster to every read from all other clusters is also computationally infeasible. Thus, we developed methods to sample representative reads from each patient and use them to create the consensus clustering. We explored a number of different methods, but two of them in particular were able to achieve significant speed improvements without significantly harming classification results. For each of them, we used a parameter "alpha", which was used to select 1 out of every alpha reads. For instance, if alpha=10, 1/10 of reads would be used.

We refer to one of our methods as "seed re-clustering" (UCLUST refers to cluster centroids as "seeds"). In this method, the reads corresponding to the cluster seeds from each patient are copied into a temporary file, and then those reads are clustered. This runs quickly, since clusters in this setting tend to have hundreds to thousands of reads, so the number of cluster seeds is relatively small. We then construct a map, which maps each seed from its original cluster number to its new cluster number from the seed clustering. We go back to the individual patient clustering output, and select 1 out of every alpha lines to be placed in the consensus clustering file. As we write the lines to the consensus file, we use the map we developed to change the cluster number for the line from the original clustering number to the new consensus cluster number. We refer to the other method as "two round clustering". We select one out of every alpha reads from the clustering output of each patient, place them in a single file, and run UCLUST on this smaller subset to arrive at a consensus clustering. The main difference between two round clustering and seed re-clustering is that here we select a distribution of reads from the original clustering to determine the consensus clustering, instead of cluster seeds/centroids.

We sought to empirically evaluate how these methods affected clustering runtime and classification accuracy, since feature vectors are generated from the clustering output and thus classification results are affected by the clustering methods. Without use of these sampling methods, UCLUST didn't finish within ARGO's 12 hour time limit on the T2D dataset, so we couldn't evaluate the classification results on that dataset and compare them to the classification results with the clustering sample methods. In order to evaluate these methods, we used the aforementioned Encephalopathy dataset, which is much smaller. We measured the time it took to perform UCLUST on all patients at once, as well as the time taken by each of our two sampling methods. We then used the respective clustering outputs to generate feature vectors and run svm-light classification (as described in the next section) and compared the results, which are discussed in the "Experimental Results" section. For UCLUST, default settings were used, except the threshold for placing two sequences in the same cluster was set to 40\% match (--id 0.40), and the --usersort option was used.

\subsection{Supervised and Multiple Instance Classification using Support Vector Machines (SVMs)}
\label{svm}

For our formulation we use SVMs to make clinical phenotype prediction for each patient sample based on their  input metagenome sequence samples represented with their clustering or taxa membership features. With our clustering approach,  we have a set of reads that have been assigned to clusters based on UCLUST and our techniques. We now generate feature vectors from each patient from this clustering output. We follow a different process of feature vector generation for svm-light and MISVM. For each method, we read through the consensus clustering output from UCLUST line-by-line, forming feature vectors as we go. UCLUST output lines provide the cluster number, the percentage match to the cluster seed, and the patient that read comes from.

For svm-light, we have one feature vector per patient. We use each cluster/OTU as a feature, meaning that the presence of various microbes are the features for the patient. Each number (feature) in the vector corresponds to a cluster/OTU, and the value for that number corresponds to how closely the read from the patient matched the cluster seed. For each input 
sequence, we use the percentage match to its cluster seed as 
the value for that feature and average the 
sequence match if multiple reads from the metagenome samples 
are assigned to the same cluster by UCLUST. Each sequence read includes a label for the patient that it came from, so each read only affects the feature vector corresponding to the patient that it came from. For example, say patient A has a read with 95\% match to OTU 2, a read that is the cluster seed of OTU 3, a read that is a 93\% match to OTU 2, a read that is a 97\% match to OTU 5, and no reads for OTU 1 or 4. Patient A's feature vector would then look like: [0 0.94 1 0 0.97]. We first initialize the patient feature vectors to all 0s and then iterate through the UCLUST output lines and update the feature vectors as we go. For MISVM, patients are bags, reads for patients are instances, and each read/instance has a feature vector. We iterate through the UCLUST output. For each read, we add that read as an instance to the relevant patient bag. The feature vector for that read contains one number: the cluster number that the read belongs to.  We compared the runtime and accuracy of the two implementations, which both used a linear kernel. In both cases, the training and test set each contained 30 patients, and each contained 15 patients with type 2 diabetes and 15 control (healthy) patients. The results are discussed in the "Experimental Results" section.

>>>>

\subsection{Experimental Setup}

The 218 publicly-available and labeled patients were downloaded from NCBI and converted to FASTQ format using the SRA toolkit, and labeled according to the labels on EBI, as described in the Materials section. Each read was assembled with SOAPdenovo2, with the k-mer length set to 51, reads cut off after 100 base pairs (original length of 180 base pairs), and the average insert size set to 350 in accordance with the reported average insert size reported in the MGWAS study \cite{qin041012}. Assembly for each file took 7-30 minutes, depending on the file size, and each patient was assembled in parallel. Combining the files into one file took 12 minutes, and clustering with UCLUST took 10 hours and 51 minutes. Without assembly first reducing the size of the data, clustering would not have been feasible. SVM-light took 10 minutes and 17 seconds to run.

\subsection{Results For Bag/Patient Labels}

Temp

\subsection{Deriving Instance "Labels"}

\begin{figure}[t]
\centering
\includegraphics[scale=0.5]{./instance-labels.png}
\caption{This diagram illustrates why static instance labels are not sufficient for phenotype prediction. A patient with 6 of the blue microbe or 6 of the green microbe may be healthy, while a patient with 3 of each is sick. Static instance labels cannot capture this relationship. This is also explored by Amores in his MIL taxonomy \cite{amores13}.} \label{instance-labels}
\end{figure}

One of the benefits of using Multiple Instance Learning methods is that we can attempt to discover instance "labels". In fact, we did not attempt to apply static, unchanging labels to individual reads or clusters, since organisms are affected by their interactions with each other. For instance, a patient with X amount of microbe A or X amount of microbe B, but not with X/2 amount of microbe A and X/2 amount of microbe B. This is illustrated in Figure \ref{instance-labels}. This example is very simplified, but it explains why static instance labels are insufficient.

However, we can infer from the SVM decision boundary which clusters appear to be most relevant to the disease diagnosis. Since feature vectors are multiplied by the weight vector of the decision boundary to determine the label of the patient, we can assume that clusters with the highest weights in the weight vector are most relevant to the disease diagnosis. For instance, if the Ith scalar in the weight vector is has the highest value of any of the weights, then cluster I is likely to play a major role in the disease. Similarly, the most negative weights in the weight vector indicate clusters whose presence in a patient indicates that they likely do not have the disease. Because the data is metagenomic, the clusters represent both phylogenetic and functional similarity, so identifying the most relevant clusters can help discover more about the pathology of the disease. For Type 2 Diabetes, which is a complex phenotype and a disease that is both common and deadly, this is potentially quite valuable.

### 2016-08-04 latex/input/intro.tex
finished related work section of paper and added diagrams; began working on results section

The human body contains one of the most dense and diverse microbial environments in the world. The human and microbial cells are collectively referred to as the \emph{human microbiome} \cite{turnbaugh_human_2007,backhed_host-bacterial_2005}. Advances in biotechnology have allowed scientists to directly interrogate the human microbiome, particularly the development of high throughput sequencing technologies, which generate massive amounts of biological data. In recent years, improving capabilities in data science have allowed for the study of \emph{metagenomics}, which involves sequencing the entire pool of microbial genomes at once. This data is usually gathered via \emph{shotgun sequencing}, which generates a number of short \emph{reads} containing bits of genomic data from the microbes of the host environment \cite{messing81}. These reads, represented as strings of nucleotides, represent only small parts of the microbe's full genome, and are not ordered in any way, which presents several challenges that will be discussed further in the paper. Metagenomics has several advantages: (i) microbes are now understood to be the underlying cause of many human diseases and are also critical to many chemical processes and overall health; (ii) it is believed that most microbes have not been laboratory-cultured and thus remain unknown; (iii) whereas other methods such as 16S rRNA analysis are mainly useful for predicting the species of microbes (phylogeny), metagenomics contains other critical information from the microbial genomes that determine how these microbes function and affect diseases and chemical processes \cite{handelsman04}. Thus, studying microbial metagenomics is an effective way to predict and model human disease, also known as clinical \emph{phenotype}.

@@

We used data from a Metagenome-Wide Association Study (MGWAS), which compares microbial metagenomic data between many patients with or without a given phenotype. This data is very large (hundreds of gigabytes) and high-dimensional (thousands of dimensions). Additionally, due to the nature of shotgun sequencing, most of the reads are not useful by themselves, and must first be assembled. \emph{Assembly} is the process of finding pairs of reads in which the end of one read overlaps with the beginning of another, signifying that they are probably contiguous reads from the same genome, and combining them. This process is repeated as much as possible to form long strings called \emph{contigs}. Assembly also reduces the size and dimensionality of the data by discarding reads that cannot be assembled successfully. Most of the discarded reads would not be particularly useful anyway, since the assembly process could not establish their relationship to other reads. The reduction in data size also allows for clustering, which is not feasible for massive datasets. Clustering uses string similarity measures to group similar reads into "clusters", which is a way of identifying which species of microbe each read corresponds to. The alternative, "aligning" the reads with known genomes, is both time-inefficient and impractical for metagenomics, since many of the involved microbes have not yet had their genomes sequenced. From the clustering output, we extract feature vectors, which are then fed into a standard Support Vector Machine (SVM) classifier.

Thus, the entire pipeline consists of assembling the reads of each patient, combining the resulting contigs from each patient into a single file, clustering the contigs, extracting features from the clustering output, and performing classification with the SVM. This process is explained in further detail in the methods section. We then compare the results of our classifier against the classifier used in the MGWAS study from which we derived our data. We show that our classifier shows significantly improved performance. This is discusses further in the Experimental Results section.

>>>>

The human body contains one of the most dense and diverse microbial environments in the world. The human and microbial cells are collectively referred to as the \emph{human microbiome} \cite{turnbaugh_human_2007,backhed_host-bacterial_2005}. Advances in biotechnology have allowed scientists to directly interrogate the human microbiome, particularly the development of high throughput sequencing technologies, which generate massive amounts of biological data. In recent years, improving capabilities in data science have allowed for the study of \emph{metagenomics}, which involves sequencing the entire pool of microbial genomes at once. This data is usually gathered via \emph{shotgun sequencing}, which generates a number of short \emph{reads} containing bits of genomic data from the microbes of the host environment \cite{messing81}. These reads, represented as strings of nucleotides, represent only small parts of the microbe's full genome, and are not ordered in any way, which presents several challenges that will be discussed further in the paper.

Metagenomics has several advantages: (i) microbes are now understood to be the underlying cause of many human diseases and are also critical to many chemical processes and overall health; (ii) it is believed that most microbes have not been laboratory-cultured and thus remain unknown; (iii) whereas other methods such as 16S rRNA analysis are mainly useful for predicting the species of microbes (\emph{phylogeny}), metagenomics contains other critical information from the microbial genomes that determine how these microbes function and affect diseases and chemical processes (\emph{functional} information) \cite{handelsman04}. Summarily, metagenomics allows us to view microbial data that is not accessible to us via traditional laboratory culturing and allows for both phylogenetic and functional profiling of those microbes. Thus, studying microbial metagenomics is an effective way to predict and model human disease, also known as clinical \emph{phenotype}.

@@

We used data from a Metagenome-Wide Association Study (MGWAS), which compares microbial metagenomic data between many patients with or without a given phenotype. This data is very large (multiple terabytes) and high-dimensional (thousands of dimensions). Additionally, due to the nature of shotgun sequencing, most of the reads are not useful by themselves, and must first be assembled. \emph{Assembly} is the process of finding pairs of reads in which the end of one read overlaps with the beginning of another, signifying that they are probably contiguous reads from the same genome, and combining them. This process is repeated as much as possible to form long strings called \emph{contigs}. Assembly also reduces the size and dimensionality of the data by discarding reads that cannot be assembled successfully. Most of the discarded reads would not be particularly useful anyway, since the assembly process could not establish their relationship to other reads. The reduction in data size also allows for clustering, which is not feasible for massive datasets. Clustering uses string similarity measures to group similar reads into "clusters", which is a way of identifying which species of microbe each read corresponds to. The alternative, "aligning" the reads with known genomes, is both time-inefficient and impractical for metagenomics, since many of the involved microbes have not yet had their genomes sequenced. From the clustering output, we extract feature vectors, which are then fed into a standard Support Vector Machine (SVM) classifier.

Thus, the entire pipeline consists of assembling the reads of each patient, combining the resulting contigs from each patient into a single file, clustering the contigs, extracting features from the clustering output, and performing classification with the SVM. This process is explained in further detail in the methods section. We then compare the results of our classifier against the classifier used in the MGWAS study from which we derived our data. We show that our classifier shows significantly improved performance. This is discussed further in the Experimental Results section.

### 2016-08-04 latex/input/materials.tex
finished related work section of paper and added diagrams; began working on results section

We used data from a well-known Metagenome-Wide Association Study by Qin et al. of Type 2 Diabetes (T2D) in Chinese patients \cite{qin041012}. The total dataset used in this study contains 368 patients \cite{qin041012}, but only 218 of them were labeled and available online. Each patient file was downloaded from NCBI\footnote{http://trace.ncbi.nlm.nih.gov/Traces/study/?acc=SRP011011}
and converted to FASTQ format using the SRA toolkit\footnote{http://www.ncbi.nlm.nih.gov/Traces/sra/sra.cgi?view=software}.
The labels were found on the EBI website\footnote{https://www.ebi.ac.uk/metagenomics/projects/SRP011011}. Each patient file was several gigabytes in size, with the total dataset size being over 600 gigabytes. Qin et al. develop a simple T2D classifier using a minimum redundancy–maximum relevance (mRMR) method \cite{peng05} for feature selection and an SVM for classification, based on the R packages "sideChannelAttack" and "e10171", respectively \cite{qin041012}. They achieved an AUC-ROC of 0.81 by training a classifier on 345 of the patients and using the remaining 23 as a test set \cite{qin041012}. While we did not have access to all of the same patients that they had, we were able to replicate their method and test it on the subset of data that was publicly available. The study also called for more extensive testing of gut microbiota classifiers \cite{qin041012}.

\subsection{Software and Hardware Details}
We used the ARGO computing cluster available at George Mason University \cite{argo}. The clustering and classification phases were run on one of the compute nodes available on the cluster. The cluster is configured with 35 Dell C8220 Compute Nodes, each with dual Intel Xeon E5-2670 (2.60GHz) 8 core CPUs, with 64 GB RAM. (Total Cores 528 and 1056 total threads, RAM$>$2TB) \cite{argo}. Source codes for 
SOAPdenovo2\footnote{SOAPdenovo2: http://soap.genomics.org.cn/soapdenovo.html} \cite{luo12}, UCLUST\footnote{UCLUST: http://www.drive5.com/uclust/downloads1\_{}2\_{}22q.html}    \cite{Edgar10}, and svm-light\footnote{svm-light: http://svmlight.joachims.org/} \cite{joachims08}
their respective websites
 and compiled on the Argo platform.

>>>>

We used data from a well-known Metagenome-Wide Association Study by Qin et al. of Type 2 Diabetes (T2D) in Chinese patients \cite{qin041012}. The total dataset used in this study contains 368 patients \cite{qin041012}, but only 218 of them were labeled and available online. Each patient file was downloaded from NCBI\footnote{http://trace.ncbi.nlm.nih.gov/Traces/study/?acc=SRP011011}
and converted to FASTQ format using the SRA toolkit\footnote{http://www.ncbi.nlm.nih.gov/Traces/sra/sra.cgi?view=software}.
The labels were found on the EBI website\footnote{https://www.ebi.ac.uk/metagenomics/projects/SRP011011}. The total size of these 218 FASTQ files was 2.3 terabytes, with an average size of 10.82 gigabytes per patient file. Qin et al. develop a simple T2D classifier using a minimum redundancy–maximum relevance (mRMR) method \cite{peng05} for feature selection and an SVM for classification, based on the R packages "sideChannelAttack" and "1071", respectively \cite{qin041012}. They achieved an AUC-ROC of 0.81 by training a classifier on 345 of the patients and using the remaining 23 as a test set \cite{qin041012}. While we did not have access to all of the same patients that they had, we were able to replicate their method and test it on the subset of data that was publicly available. The study also called for more extensive testing of gut microbiota classifiers \cite{qin041012}.

\subsection{Software and Hardware Details}
We used the ARGO computing cluster available at George Mason University\footnote{http://orc.gmu.edu/research-computing/argo-cluster/argo-hardware-specs/}. The clustering and classification phases were run on one of the compute nodes available on the cluster. The cluster is configured with 35 Dell C8220 Compute Nodes, each with dual Intel Xeon E5-2670 (2.60GHz) 8 core CPUs, with 64 GB RAM. (Total Cores 528 and 1056 total threads, RAM$>$2TB). Source codes for 
SOAPdenovo2\footnote{SOAPdenovo2: http://soap.genomics.org.cn/soapdenovo.html} \cite{luo12}, UCLUST\footnote{UCLUST: http://www.drive5.com/uclust/downloads1\_{}2\_{}22q.html}    \cite{Edgar10}, and svm-light\footnote{svm-light: http://svmlight.joachims.org/} \cite{joachims08}
their respective websites and compiled on the Argo platform.

### 2016-08-04 latex/input/related.tex
finished related work section of paper and added diagrams; began working on results section

The \emph{assembly} problem involves combining overlapping short reads into longer sequences called \emph{contigs}.

@@

The \emph{clustering} problem in this context  involves grouping input short reads  such that reads within a group are 
similar to each other. This process may lead to groups that are organism-specific and is 
unsupervised in nature. 
%
The clustering process does not attempt to provide an automated labelling of the input reads. 
%
The clusters/groups obtained from an input metagenome sample 
are referred to as  Operational Taxonomic Units (OTUs), and the number of 
OTUs gives an approximation of species diversity in a sample \cite{schloss2009introducing,schloss2005introducing,sun2009esprit}.
%
These  approaches are not constrained due to the absence of a complete coverage in taxonomic databases. Some 
environmental samples contain microbial organisms that have never been cultured in a laboratory, and thus those organisms do not exist in genomic databases.
As such,  clustering of 
sequence reads has several advantages: (i)
        it can lead to an improved  metagenome assembly, (ii) it can be used 
        for computing species diversity metrics \cite{bibm2012} 
        and (iii)
        the  reduced computational
        complexity within several work-flows that analyze only
        cluster representatives, instead of individual sequences
        within a sample.

CD-HIT \cite{Li01072006}, UCLUST \cite{Edgar10}, CROP \cite{Hao01032011}, MC-MinH \cite{sdm2013a} and 
MC-LSH \cite{bibm2012}
are some of the popular 
metagenome sequence clustering approaches used for binning. 
%
UCLUST, MC-MinH and MC-LSH  are
greedy approaches that achieve computational efficiency 
by using either hash-based indexing or matching of gap-less sequences called seeds (instead 
of expensive sequence alignment) and followup with an incremental clustering approach 
that does not involve comparing all pairs of input sequences. We use UCLUST within our study, which is one of the most widely used and cited metagenome clustering methods and has been shown to be amongst the most effective in terms of speed and accuracy in benchmarking studies \cite{bonder090112, sun042711}. 
%

UCLUST \cite{Edgar10} follows a greedy, iterative clustering approach. As a first 
step, this approach identifies exact matches of fixed length between sequence pairs known 
as seeds. These seeds are then extended by allowing for a few mismatches and/or gaps between 
the aligned pairs. Seeds scoring above a certain threshold are chosen as high segment pairs (HSPs)  and used for further processing. As such, the step eliminates a lot of pairwise comparisons. Then UCLUST follows an incremental approach for assigning the input sequences to the different bins. The clustering solution is initialized as an empty list. Sequences are then incrementally added to the clusters existing within the list. Each unassigned input sequence is compared to the cluster representatives within the list using the fast indexing search technique that uses the HSPs. If a match is found with one of the cluster representatives, then that sequence is assigned to that particular cluster; otherwise, the input sequence forms a new cluster. Matches are computed by comparing two reads, character by character, and dividing the number of characters at the same position in both reads that match by the length of the longer of two reads. UCLUST ensures that the cluster representatives are sequences with the largest length.

>>>>

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{./de-bruijn.png}
\caption{This diagram illustrates a basic de Bruijn graph, as explained in the text. Notice the double arrow from TAT to ATT, allowing a path to go through that route twice. Thus, the longest contig that can be created by following a path is GTATTATTGGA, which is what we want.} \label{de-bruijn}
\end{figure}

The \emph{assembly} problem involves combining overlapping short reads (usually less than 1000 base pairs) into longer sequences called \emph{contigs} (often tens of thousands of base pairs). For instance, if one read ends with the same relatively large nucleotide string that another read starts with, the reads are likely to be overlapping fragments from the same genome, and can thus be combined into one contig. This can be done either \emph{de novo} (in an unsupervised manner) or by referencing sequences against known contigs. We focus on de novo assembly, because we wish to keep our pipeline as unsupervised as possible.

One of the main purposes of assembly is to determine the whole genomes of microbial species, the vast majority of which have not or cannot be laboratory cultured, from sequencing reads \cite{zerbino08}. Even if full genomes cannot be assembled, combining reads into larger contigs can still make them much more useful for clustering and classifiers, because the contigs will contain more phylogenetic and functional information than short reads. This is because short reads of less than 1000 base pairs constitute only a tiny fraction of microbial genomes, which are usually hundreds of thousands to millions of base pairs, making it difficult to ascertain much about the phylogeny of individual reads. Many modern sequence reads are produced by Next-Generation and High-Throughput Sequencers, which usually produce these short reads. Metagenomics additionally poses its own set of challenges, due to very large datasets and lack of knowledge about how many species are present and in what relative abundances \cite{namiki12}. Thus, metagenome assembly is a relatively new and challenging field.

Some of the popular genome and metagenome assembly approaches include SOAPdenovo2 \cite{luo12}, IDBA-UD \cite{peng12}, Velvet \cite{zerbino08}, MetaVelvet \cite{namiki12}, ABySS \cite{simpson09}, and Ray Meta \cite{boisvert12}. We used SOAPdenovo2 in our study, because it was the assembler used in the original MGWAS study \cite{qin041012} and because it has been shown to be one of the fastest assembly algorithms \cite{peng12}. SOAPdenovo2 first constructs a type of directed graph called a \emph{de Bruijn} graph that represents the overlaps between different sequences \cite{li10}. Reads are divided into strings of length K called \emph{k-mers}; these k-mers are the nodes of the graph \cite{zerbino08}. The choice of K is up to the user, and is important for having good assembly results. The k-mer nodes in the graph have an edge between them if a read contains those k-mers in order with an overlap of K-1 nucleotides, and the direction of the edge indicates in which order the k-mers appear \cite{zerbino08}. SOAPdenovo2 then cleans this graph by removing nodes/sequences with few or no connections with other sequences, eliminating "tips" that represent likely sequencing machine errors, and removing redundant edges \cite{li10}. The contigs are then formed by combining reads according to the de Bruijn graph: each contig represents a directed path in the graph \cite{zerbino08}. There are variations on how exactly de Bruijn graphs are represented and implemented, but Figure \ref{de-bruijn} illustrates the general process.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{./uclust-similarity.png}
\caption{This diagram illustrates the similarity computation between two sequences in UCLUST. Terminal characters are excluded, so the length of the reads is 10. In the first 10 characters of each string, there are 5 matching characters that are in the same position in the string. Thus, the similarity is 5/10 = 50\%.} \label{uclust-similarity}
\end{figure}

The \emph{clustering} problem in this context  involves grouping input short reads  such that reads within a group are similar to each other. The clusters obtained from this process are referred to as Operational Taxonomic Units (OTUs). OTUs represent a group of equivalent or similar organisms. Accordingly, the number of OTUs in a sample gives an approximation of the species diversity in that sample  \cite{schloss2009introducing,schloss2005introducing,sun2009esprit}.
In addition to approximating species diversity, clustering has several other key advantages. Because clustering is always de novo (unsupervised), it is not limited by the species that are covered in taxonomic databases. This detail is important, because it is believed that most microorganisms that reside in the human body have not been laboratory cultured \cite{handelsman04}. Clustering also reduces computational costs by allowing analyses to operate on entire clusters instead of on each read. Finally, clustering helps the classification process by allowing feature vectors to be built at the OTU level, instead of using individual short reads that contain little biological information.

UCLUST \cite{Edgar10}, CD-HIT \cite{Li01072006}, mothur \cite{schloss2009introducing}, DOTUR \cite{schloss2005introducing}, CROP \cite{Hao01032011}, MC-MinH \cite{sdm2013a}, and 
MC-LSH \cite{bibm2012} are some of the popular  nucleotide sequence clustering approaches. We use UCLUST within our study, which is one of the most widely used and cited metagenome clustering methods and has been shown to be amongst the most effective in terms of speed and accuracy in benchmarking studies \cite{bonder090112, sun042711}. 
%

UCLUST seeks to ensure that, for some similarity T, the following conditions hold: (i) all cluster centroids have a similarity of less than T to each other; and (ii) all points in a cluster have a similarity of greater than T to the cluster centroid \cite{Edgar10}. Thus, each \emph{centroid} defines the center of a cluster, and the distance T defines the radius of the cluster, such that any point that has a similarity of greater than T to the centroid is within the radius and is thus part of the cluster. UCLUST is a heuristic algorithm that has several optimizations to improve speed, thus condition (i) above is not always guaranteed \cite{Edgar10}. UCLUST proceeds in a greedy, iterative manner. The first sequence in the input file becomes a new cluster centroid. For each new sequence in the file, it is compared with each of the existing cluster centroids in order. As soon as it is compared with a centroid that it has a similarity of greater than T with, it becomes part of that cluster. If the read is not similar enough with any of the existing cluster centroids, it becomes the centroid of a new cluster. The similarity measure T is defined as a string similarity between the two nucleotide sequences that counts the number of character placements that they have in common and then divides that number by the length of the reads, with terminal characters excluded \cite{Edgar10}. This is illustrated in figure \ref{uclust-similarity}.
